---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data

We extract the .csv file from the zip and adjust the column "date" withe a Date format:
```{r}
DF <- read.csv(unz("activity.zip", "activity.csv"), colClasses=c("integer", "Date", "integer"))

head(DF)
summary(DF)
str(DF)
````
## What is mean total number of steps taken per day?

First we select the data we need and group it per date. Then we sum the steps for each day:

````{r, message=FALSE, warning=FALSE}
library(dplyr)
stepsDay<-DF %>%
        select(steps, date) %>%
        group_by(date) %>%
        summarise(sumSteps=sum(steps))

head(stepsDay)
````

Now we prepare the histogram of the total number of steps taken each day:
(we adjust the bindwith to 1.000 steps)

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
ggplot(data=stepsDay, aes(sumSteps)) + 
        geom_histogram(binwidth = 1000) +
        labs(title="Steps taken each day", 
             x = "Number of steps each day", y = "Number of times in a day")
```

Now we calculate the **mean** and **median** total number of steps taken per day with this code 
```{r, message=FALSE, warning=FALSE}
mean(stepsDay$sumSteps, na.rm=TRUE)
median(stepsDay$sumSteps, na.rm=TRUE)
```

## What is the average daily activity pattern?

Now we make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):

```{r, message=FALSE, warning=FALSE}
#To plot average daily activity
DF %>%  group_by(interval) %>% 
        summarize(mean=mean(steps, na.rm=TRUE)) %>% 
        plot(., type="l", main="Average daily activity", 
             xlab="5-minute interval", ylab="Average steps across all days")

#To display the interval with the max. number of steps in average on the plot
max<-DF %>%  group_by(interval) %>% 
        summarize(mean=mean(steps, na.rm=TRUE)) %>% 
        slice(which.max(mean)) %>% select(interval)

abline(v=max, col="purple")
axis(1, at=max, labels = max, pos=0, col.axis="purple")
```

In purple we see the interval with the maximum number of steps. And here the exact  
interval and his mean:
```{r, message=FALSE, warning=FALSE}
DF %>%  group_by(interval) %>% 
        summarize(mean=mean(steps, na.rm=TRUE)) %>% 
        slice(which.max(mean)) #slice selects rows by position
```



## Imputing missing values
Note that there are a number of days/intervals where there are missing
values (coded as `NA`). The presence of missing days may introduce
bias into some calculations or summaries of the data.

````{r}
summary(DF)

````
1. There are **`r sum(is.na(DF))`**  missing values in the dataset


2. We will fill the NA's with the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
````{r}
# Split main DF in two DF, one with NAs and one without NAs
DF_noNA <-DF %>% filter(complete.cases(.))
DF_NA <-DF %>% filter(!complete.cases(.))

# Create DF with intervals and their step means
meanSteps <- DF %>% group_by(interval) %>% summarize(mean=mean(steps, na.rm=TRUE))

# Merge means with DF with NAs by "interval"
DF_NA <- merge(meanSteps, DF_NA, by = "interval", all.y=TRUE)
DF_NA <- DF_NA[,c(2,4,1)] #remove column not needed and reorder columns
colnames(DF_NA) <-c("steps", "date", "interval") #rename columns

# Merge DF with no NA's and with the DF with means
DF_new <- rbind(DF_noNA, DF_NA)

````


Here the histogram of the new DF

````{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
stepsDay_new<- DF_new %>%
        select(steps, date) %>%
        group_by(date) %>%
        summarise(sumSteps=sum(steps)) 


ggplot(data=stepsDay_new, aes(sumSteps)) + 
        geom_histogram(binwidth = 1000) +
        labs(title="Steps taken each day (new DF)", 
             x = "Number of steps each day", y = "Number of times in a day")

```
  
Here the **mean** and **median** of the new DF 

```{r, message=FALSE, warning=FALSE}
mean(stepsDay_new$sumSteps, na.rm=TRUE)
median(stepsDay_new$sumSteps, na.rm=TRUE)

summary(stepsDay$sumSteps) # before

summary(stepsDay_new$sumSteps) # now
```

Now the estimates have changed as follows:  
-  We have now more observations as shown in the histogram (we added data to the NAs)  
-  Mean stays the same (the strategy was to use the mean to fulfil the missing values)  
-  Now the median is the same as the mean  



## Are there differences in activity patterns between weekdays and weekends?
For this part we are going to use the data set with the filled-in missing values.

We will add a new column with the information, Weekday or Weekend.

````{r, message=FALSE, warning=FALSE}
library(dplyr)

DF_new$weekday <- as.factor(
        ifelse(weekdays(DF_new$date) %in% c("Samstag","Sonntag"), "Weekend", "Weekday")
        ) 

str(DF_new)
````

Now we make a plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

````{r, message=FALSE, warning=FALSE}

library(dplyr)
library(lattice)
DF_new_weekday <- DF_new %>% 
        group_by(interval, weekday) %>%
        summarise(steps = mean( steps, na.rm=TRUE ))

xyplot(steps~interval | weekday, data = DF_new_weekday,
       type="l", ylab="Number of steps", 
       xlab="Interval",
       layout=c(1,2))

````
